---
title: "plantR Tutorial"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        toc: true
        number_sections: true
    md_document:
      variant: gfm
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{plantR tutorial}
  %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, message= TRUE}
#aqui carrega o pacote que está no seu HD, não precisa instalar ele para trabalhar
devtools::load_all()
```


# Introduction

# Data entry  (FUNCTIONS NOT CREATED YET)

## List of collections available for download
	
## Obtaining the links for the DwC, EML and RTF files and other collection metadata
	
## Downloading the content for each collection
	
### speciesLink
		
### Jabot
		
### GBIF
    
### SiBBr
		
## Filtering of the downloaded data? By names or other fields...


# Data processing

## Names, collector number and dates: function `formatOcc()`


```{r, eval = FALSE}
occs <- example
occs.clean <- fixField(occs, origin = "jabot")
occs.clean2 <- formatOcc(occs.clean)
```


## Editing localities and match with gazetteer information: function `formatLoc()`
	
## Editing coordinates and defining working coordinates: function `formatCoord()` (TO BE CHECKED)


# Data validation

## Taxonomic validation: function `validateTax()`
	
## Geographical validation: function `validateCoord()`
	

# Duplicated specimens (FUNCTIONS NOT CREATED YET)

## Prepare to merge: function `dupPrep()`
	
## Find duplicates across collections: function `dupSearch()`
	
## Homogenizing information within duplicates: function `dupInfo()`
	

# Data description  (NEED TO CHECK)

## Number of collections, specimens and duplicates retrieved: 
	
