---
title: "plantR Tutorial"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        toc: true
        number_sections: true
    md_document:
      variant: gfm
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{plantR tutorial}
  %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

This tutorial provides basic explanations on the functions related to data download, processing and validation using `plantR`. For each of these steps, there is a more specific tutorial presenting all functionalities more in detail.

Let's first install and load the package from [CRAN](https://CRAN.R-project.org) with:

```{r, eval = FALSE}
install.packages("plantR")
library(plantR)
```

or from [GitHub](https://github.com) with:
```{r, eval = FALSE}
library("devtools")
install_github("LimaRAF/plantR")
library("plantR")
```


# Data entry

You can download species occurrence data directly from R, using the functions provided by `plantR` or by other R packages. For this tutorial, we are going to use the `plantR` function `rspeciesLink`, which downloads data from [CRIA](http://www.cria.org.br/) (Centro de Referência em Informação Ambiental). For a detailed tutorial of how to download data from different sources, check a more detailed tutorial [here](#INCLUDE HERE LINK TO SARA'S DETAILED TUTORIAL). 

You can also use .csv files containing personal data or data downloaded manually from online data repositories, as long as the the column names match those accepted by `plantR` (e.g. DarwinCore). See function `fixField()` below for more details on standandard columns names.
	
## Downloading occurrence data with `plantR`

We will use as an example a dataset downloaded from speciesLink, using the function `rspeciesLink`, which allows a variety of ways on how to search for data. In this tutorial, we will use data for (one/some) common Neotropical tree(s):

```{r, eval = FALSE}
#spp <- c("Trema micrantha",
#        "Myrcia splendens",
#        "Casearia sylvestris")
spp <- c("Trema micrantha")
```

After defining the species name we want to obtain occurrence data, we proceed to the download itself. Here, we will use only herbarium plant specimens (i.e. vouchers) for the names listed above, iincluding the list of synonyms accepted in the Brazlian Flora 2020 project (this may take a few minutes):

```{r, eval = FALSE}
example <- rspeciesLink(scientificName =  spp,
                     Scope = "plants",
                     basisOfRecord = "PreservedSpecimen",
                     Synonyms = "flora2020")
```

The occurrence data is now stored in the object `example`. We can inspect how many occurrences were downloaded (about 4500 records and over 40 fields):

```{r, eval = FALSE}
dim(example$data)
```

# Data processing

## Preparing the input data: function `fixField()`

Before we can start to process the input data, we need to be sure that our input data has all the required fields and that they are in the required format.

We use the function `fixField()` to set up the require and optional fields and to drop fields without essential information for the data processing and validation. We save the output of the function in an object with a different name ('occs').

```{r, eval = FALSE}
occs <- fixField(example$data, origin = "splink")
```

Note that the function returns some warning meassages. It is **VERY** important to carefully understand these warnings before we can continue. 

The first warning says that the name of the column header 'scientifname', provided in the input data, was replaced by the standard name 'scientificName', the standard name used by `plantR`. The second warning also reports that a header name was replaced by a standard header name, 'occurrenceID' in this case. Third warning is less important because it relates to optional fields. The last warning is more important, since they may indicate typos in the names of important hearders.

## Standardizing names, collector number and dates: function `formatOcc()`

Particularly when working with data coming from multiple secondary sources, it is important to standardize the notation used by different collections and sometimes by different collectors within the same collection.

We first create the new columns that will receive the edited fields.

```{r, eval = FALSE}
occs$recordedBy.new = occs$recordedBy
occs$recordNumber.new = occs$recordNumber
occs$identifiedBy.new = occs$identifiedBy
occs$year.new = occs$year
occs$yearIdentified.new = occs$yearIdentified
```

The year of collection is sometimes stored on the field 'eventDate' and not on the field 'year'. In this case, we should replace all missing information from the field 'year' by the informations stored in the field 'eventDate'. Since this is not the case here, we procees to the next data processing step.

We then prepare the new fields we created using functions `fixName()`, `colNumber()` and `getYear()`. We chose to remove all species characters from authors and identicador names (argument `special.char = FALSE`), which is the indicated option for further data processing and validation. We also set the characters that will represent occurrences without number of collection ("s.n.") and without the year of collection and identification ("n.d.").  

```{r, eval = FALSE}
occs$recordedBy.new = unlist(lapply(occs$recordedBy.new, fixName, special.char = FALSE))
occs$recordNumber.new = unlist(lapply(occs$recordNumber.new, colNumber, noNumb = "s.n."))
occs$identifiedBy.new = unlist(lapply(occs$identifiedBy.new, fixName, special.char = FALSE))
occs$year.new = unlist(lapply(occs$year.new, getYear, noYear = "n.d."))
occs$yearIdentified.new = unlist(lapply(occs$yearIdentified.new, getYear, noYear = "n.d."))
```

Next, we format the names of the collector and identificator (function `formatName()`)
```{r, eval = FALSE}
occs$recordedBy.new = unlist(lapply(occs$recordedBy.new, formatName))
occs$identifiedBy.new = unlist(lapply(occs$identifiedBy.new, formatName))
```

We then separate first and auxiliary names for multiple names and then convert them to the TDWG format (function `tdwgNames()`). To do so, we set the argument 'out' of function `tdwgNames()` to "aux" and "first" (by default `tdwgNames()` edits and returns all names from multiple name strings).

We also define that the symbol that will separate different names, which in our example will be a semi-colon followed by a space.

```{r, eval = FALSE}
occs$recordedBy.aux = unlist(lapply(occs$recordedBy.new, tdwgNames, out = "aux", sep.out = "; "))
occs$identifiedBy.aux = unlist(lapply(occs$identifiedBy.new, tdwgNames, out = "aux", sep.out = "; "))
occs$recordedBy.new = unlist(lapply(occs$recordedBy.new, tdwgNames, out = "first"))
occs$identifiedBy.new = unlist(lapply(occs$identifiedBy.new, tdwgNames, out = "first"))
```

We can inspect what is the result of this separation between first and auxiliary names:
```{r, eval = FALSE}
head(occs[,c("recordedBy","recordedBy.new","recordedBy.aux")], 3)
```

It is also useful for the validation process to standardize the notation for missing collector and identificator names. To do so, we use the function `missName()` and again we set missing names as "s.n.".

```{r, eval = FALSE}
occs$recordedBy.new = missName(occs$recordedBy.new, type = "collector", noName = "s.n.")
occs$identifiedBy.new = missName(occs$identifiedBy.new, type = "identificator", noName = "s.n.")
```

For some of the operations we will perform, it is also useful to extract the last name of the first collector, that is stored in our formated field 'recordedBy.new'.

```{r, eval = FALSE}
occs$last.name = unlist(lapply(occs$recordedBy.new, lastName, noName = "s.n."))
```

We can inspect what these functions are doing by comparing the original and edit columns (only the first 15 records)

```{r, eval = FALSE}
head(cbind(occs$recordedBy, occs$recordedBy.new, occs$recordedBy.aux), n = 15)
head(cbind(occs$recordedBy, occs$last.name), n = 15)
head(cbind(occs$identifiedBy, occs$identifiedBy.new), n = 15)
head(cbind(occs$year, occs$eventDate, occs$year.new), n = 15)
head(cbind(occs$yearIdentified, occs$yearIdentified.new), n = 15)
```

These functions handle well most formats but not all of them  (e.g. name format 'A. Custódio, Filho' or date format "Jul-02"). However, this standardization will be very useful for the search of duplicated specimens across herbaria (see below).

All the previous steps are important to understand the edtiton of each field. But, `plantR` can execute all these steps at once using the wrapper function `formatOcc()`:

```{r, eval = FALSE}
occs1 = fixField(example$data, origin = "splink")
occs1 = formatOcc(occs1)
```

Let's compare if the output of the two approaches are the same, column by column:

```{r, eval = FALSE}
table(occs$recordedBy.new == occs1$recordedBy.new)
table(occs$identifiedBy.new == occs1$identifiedBy.new)
table(occs$year.new == occs1$year.new)
table(occs$yearIdentified.new == occs1$yearIdentified.new)
table(occs$last.name == occs1$last.name)
rm(occs1)
```

Yes! So, the new data frame created contain the original and edited/formatted information regarding the name of the collector and of the identificator, the dates of collection and identification, and number of collection. 


## Editing locality information: function `formatLoc()`

One important step to find missing geographical coordinates and to validate those coordinates provided with the occurrence is to standardize the fields containing locatity information, namely "country", "stateProvince", "municipality" and "locality".

`plantR` have a function to edit and standardize locality fields: function `fixLoc()`. By default, this function formats all four locality fields simultaneously (returns the entry data.frame with the new columns), but the user can chose one field at a time by changing the function argument 'adm.levels' (returns a vector). However, some of the editing processes become more complete if all the information is available for the four fields.

To exemplify, we will use all locality fields:

```{r, eval = FALSE}
locals = fixLoc(occs, adm.levels = c("country","stateProvince","municipality","locality"), scrap = TRUE)
locals1 = fixLoc(occs, adm.levels = c("country","stateProvince","municipality","locality"), scrap = FALSE)
```
Note that we performed two different ways to edit the locality fields. The argument 'scrap' controls the search for missing municipality information from the field locality. It also performs some extra editing and croping of the locality field in order to obtain more standardized locality descriptions. We can compare the original and the edited locality fields, by inspecting the number of records in each class of resolution: 
	
```{r, eval = FALSE}	
table(locals$resol.orig, useNA = "always")
table(locals1$resol.orig, useNA = "always")
```

If the country, state and municipality fields are given, they remain unaltered. Only missing information in this fields are completed by the editing and scrapping process.

Although the search for missing localities results in no NAs, it does not mean that the information retrieved is valid. It depends if the information is actually available in the locality field and how it is arranged. To make sure that the missing information obtained is indeed valid, we need to cross it with a gazetteer. If the municipality exists for a given state and country, then it is deemed as being a valid municipality, which will be further used on the coordinate validation process. If it does not exist, then a higher administrative level is used as the reference 'resolution' of the locality information. 

```{r, eval = FALSE}	
locals  = getLoc(locals, gazet = "plantR")
locals1 = getLoc(locals1, gazet = "plantR")
```	

Note that the edited version of the localities is different from the original localities regarding their resolution. Many information provided at locality level was only retrieved at municipality level, which is expected due to the completeness of the gazetteer used (not all possible localities are stored in the standard `plantR` gazetteer). But many of them were only retrieved at state/province level, meaning that fields cointainig state and municipality information may need to be checked. 

Some of the localities were retrieved in the default gazetteer provided with plantR, which is more complete for Latin America and has more detailed information at locality level mainly for Brazil. You can use the argument `gazet` to use your personal gazetteer instead of `plantR` default, as long as it contains the same fields.

Finally, we merge the data frame with the previous data processing (function `formatOcc()`) with the output of the locality data processing:

```{r, eval = FALSE}	
occs  = cbind.data.frame(occs, locals[,c("loc.correct","latitude.gazetteer","longitude.gazetteer","resolution.gazetteer","loc.check")])
```	

Besides the coordinates of all localities at the best locality resolution available, these procedure also performs a initial checking of the locality names provided, which is stored in the column 'loc.check': 

```{r, eval = FALSE}	
table(occs$loc.check)
```	

In this column, specimens for which the locality information provided could not be found in the gazetteer are flagged with a "check_" and thus they had their resolution downgraded. We also flag all the specimens that retained their resoltution (i.e. "ok_same") and those that could be validate at a better locality resolution, flagged with a "ok_" (e.g. 'ok_state2municipality': 'stateProvince' level that is now at the 'municipality' level).

## Editing original coordinates and defining working coordinates: function `formatCoord()`

Previous to the validation of the orginal geographical coordinates, it is important to make sure that they are given and if they are in the required format: non-zero, non-NA [decimal degrees](https://en.wikipedia.org/wiki/Decimal_degrees), with decimal digits separated by points.

```{r, eval = FALSE}	
occs = prepCoord(occs)
```	

Geographical coordinates are often missing or are provided in a format that the function `prepCoord` cannot convert to decimal degrees. For the dataset we are using in this tutorial, we can inspect the proportion of occurrences without geographical coordinates: 

```{r, eval = FALSE}	
table(!is.na(occs$decimalLatitude.new))/dim(occs)[1]
```	

As we can see, about 20% of the occurrences records have no coordinates. Therefore, one may want to use the coordinates obtained from the gazetteer to replace the missing coordinates. In `plantR` we use the function `workCoord` to do this procedure. This function also flags the origin and the probable resolution of the original coordinates: 

```{r, eval = FALSE}	
occs = workCoord(occs)
table(!is.na(occs$decimalLatitude.new))/dim(occs)[1]
```	

Now, there is a geographical coordinate at the best resolution available for almost all occurrences.

# Data validation

## Taxonomic validation: function `validateTax()`

### Species names spelling, format and synonyms 


### Confidence level of species determinations

	
## Geographical validation: function `validateCoord()`
	

# Duplicated specimens (FUNCTIONS NOT CREATED YET)

## Prepare to merge: function `dupPrep()`
	
## Find duplicates across collections: function `dupSearch()`
	
## Homogenizing information within duplicates: function `dupInfo()`
	

# Data description  (NEED TO CHECK)

## Number of collections, specimens and duplicates retrieved: 
	
