---
title: "An Introduction to __plantR__"

date: "`r format(Sys.time(), '%d %B %Y')`"
author: 
  - Renato A. F. de Lima^[Naturalis Biodiversity Center and Universidade de São Paulo, https://github.com/LimaRAF], Sara R. Mortara^[Jardim Botânico do Rio de Janeiro, https://github.com/saramortara] and Andrea Sánchez-Tapia^[Jardim Botânico do Rio de Janeiro,https://github.com/AndreaSanchezTapia]
output:
    rmarkdown::html_vignette:
        toc: true
        number_sections: true
    md_document:
      variant: gfm
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{An Introduction to __plantR__}
  %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # for development
```

```{r load, echo = FALSE, eval = TRUE, include=FALSE}
devtools::load_all() # for development
```


<br/><br/>

# Introduction

The databases of biological collections are becoming increasingly available
online, providing an unprecedented amount of species records for
biodiversity-related studies. Managing the information associated with species
records is an important but difficult task. The notation of collectors' names,
numbers and dates vary between collections, and sometimes within them. In
addition, it is often difficult to validate the localities, geographical
coordinates and identifications associated with individual species records,
especially when working with thousands or millions of them. Thus, having tools
to process and validate large amounts of records can be quite handy.

__plantR__ is an R package that was developed to manage, standardize and
validate the information associated with species records from biological
collections (e.g. herbaria). It can be used for data coming from a single
collection or different biodiversity databases, such as
[GBIF](https://www.gbif.org/). Moreover, __plantR__ can be used by collection
curators to manage their databases and  by final users of species records (e.g.
taxonomists, ecologists and conservationists), allowing the comparison of data
across collections.

## Main features and workflow

The package __plantR__ provides tools to standardize the information from
typical fields associated with species records, such as collectors' and species
names. In addition, __plantR__ proposes a comprehensive and reproducible
workflow to apply those tools while handling records from biological collections,
which includes the following steps:

1. the download of species records for a list of species names, collections
codes or other search fields;

2. the batch standardization of typical fields (e.g., collector name);

3. the validation of the locality and geographical coordinates of the records,
based on maps and gazetteers;
  
4. the spell-checking and validation of botanical families and species names
using different taxonomic backbones (e.g. Flora do Brasil);

5. the assessment of the confidence level of species identifications, based on a
global list of plant taxonomists.
    
6. the retrieval of duplicated specimens across collections, including the
homogenization of the information within duplicates.

7. the summary of species data and validation steps, and the (fast) export of
the validated records by groups (e.g. families or countries).
    
## Basic assumptions and limitations

The tools provided by __plantR__ do not edit the columns with the original
information. All the outputs of each editing or validation step are stored in
new, separated columns added to the original data. This is important for the
collection curation process because it allows the comparison between the
original and edited information. However, it increases the number of columns in
the dataset, which may become a problem while managing and saving big datasets.

__plantR__ was initially developed to manage plant records from herbaria.
Therefore, some of the tools offered by the package are exclusive to plants,
particularly the checking of species names. However, some of its main features
are expected to work for other groups of organisms as well, as long as the data
structure is similar.

Currently, the download of records is available for the [Global Biodiversity
Information Facility (GBIF)](https://www.gbif.org/), and
[speciesLink](http://splink.cria.org.br/), but the user can also provide his own
dataset as an input. Future versions of the package may include the download
from data stored in [JABOT](http://jabot.jbrj.gov.br//).

Name editing and standardization cover most of the typical variation in the
notation of people's names, trying to provide standardized outputs in the TDWG
format. The same applies to collection codes, collector numbers and dates.
However, __plantR__ does not handle all possibilities of notation. So, some
double-checking and corrections may be needed depending on the user's goals.

Regarding the validation of geographical coordinates. In the case of invalid or
missing coordinates, we assume that the locality information associated with the
record (e.g. country, state, county) is correct (i.e. locality prevails over
coordinates), and so working coordinates are taken from a gazetteer. It is
important to note that if the locality information is indeed mistaken (eg. wrong
county name), then even if the original coordinates are good, they will not be
validated (record locality and coordinate locality don't match) and may be
replaced by coordinates taken from the gazetteer.

Currently, geographical validation can be performed at the county level for
Latin American countries and at the country level for the rest of the world. We
provide a gazetteer to retrieve and check localities and geographical
coordinates, which is currently biased towards Latin American countries,
particularly Brazil. Therefore, the validation of geographical coordinates
provided by other R packages (e.g. __CoordinateCleaner__) may be more
appropriate for studies extending outside Latin America.

Taxonomical validation is performed based on (i) the correction of plant family
and species names (i.e. synonyms, typos) and (ii) the confidence level on the
species identification, based on a dictionary of plant taxonomist names from all
over the world. For i, names are currently checked against the [Flora do Brasil
project](http://floradobrasil.jbrj.gov.br/reflora/listaBrasil/ConsultaPublicaUC/ConsultaPublicaUC.do)
and [The Plant List](http://www.theplantlist.org/), using the R packages
__flora__ and __Taxonstand__. Future versions may include comparisons against
Tropicos and GBIF backbones.

During the assessment of the taxonomic confidence level of the identifications,
we did not attempt to set priorities for different specialists within a given
family. That is, all species names determined by a specialist within their
family of expertise are taken as being correct. Although we recognize that
there are specialists for genera within a family, the validation process is
currently performed only at the family level. In the case of conflicting species
identification among family specialists for duplicates across collections, we
assume the most recent identification as being the valid one.

__plantR__ provide tools for searching for duplicated records across collections.
This search makes more sense when data from different collections are combined
and it performs welleven when using relatively large datasets (i.e. millions of records).
However, the retrieval of duplicates greatly depends on the completeness of the
input information, the notation standards and if __plantR__ is able to handle
those differences in notation across collections. In addition, true duplicates
may not be found due to typos and false duplicates may be returned if the
duplicate search fields are too flexible.


<br/><br/>

# Using __plantR__


## Installation
The package can be installed and loaded from [GitHub](https://github.com) with:
```{r, eval = FALSE}
library("remotes")
install_github("LimaRAF/plantR")
library("plantR")
```

## Main features

### Data entry

Users can provide their own dataset or download data using one of __plantR__
download functions. They include the function `rspeciesLink()`:

```{r splink-download, eval = TRUE, collapse = TRUE}
occs_splink <- rspeciesLink(species = "Euterpe edulis")
```

This function can also be used to search from records based on localities,
collections, and other options (see ?rspeciesLink for details).

__plantR__ also provides the function `rgbif2()`, which is a wrapper of the
function `rgbif()` of the __rgbif__ package, with an standardized output:

```{r gbif-download, eval = TRUE, collapse = TRUE}
occs_gbif <- rgbif2(species = "Euterpe edulis")
```

```{r load-data, eval=FALSE, echo=FALSE}
occs_splink <- read.csv("results/speciesLink.csv")
occs_gbif <- read.csv("results/gbif.csv")
```

#### Field names

It is important to make sure that the field names of the input data follow the
DarwinCore format. In __plantR__ this is performed using the function
`formatDwc()`, which joins data from different sources (e.g. GBIF and
speciesLink) and standardizes their field names:

```{r bind-data, eval = TRUE, collapse = TRUE}
occs <- formatDwc(splink_data = occs_splink, 
                  gbif_data = occs_gbif)
```


### Data editing


#### Collection codes, people names, collector number and dates

The names of the collections, collectors and identifiers, as well as the
collection numbers and dates can be edited using the function `formatOcc()`:

```{r formatOcc, eval = TRUE, collapse = TRUE}
occs <- formatOcc(occs)
```

#### Locality information

The locality information associated with the occurrence data (e.g. country or
city names) can be standardized using the function `formatLoc()`:

```{r formatLoc, eval = TRUE}
occs <- formatLoc(occs)
```

#### Geographical coordinates

The geographical coordinates are prepared using function `formatCoord()`, which
guarantees that they are in a good format for validation (i.e. decimal degrees).
This function also retrieves missing coordinates from a gazetteer based on the
locality information:

```{r formatCoord, eval = TRUE, collapse = TRUE}
occs <- formatCoord(occs)
```

#### Species and family names

In this example, although we have dowloaded data for a single species (i.e.
*Euterpe edulis* Mart.), there are differences in the notation of botanical
family and species names, some of them being synonyms. To obtain only valid
names, we use the function `formatTax()`:

```{r formatTax, eval = TRUE, collapse = FALSE}
occs <- formatTax(occs)
```

### Data validation

#### Locality information

Once the new columns with the edited and standardized information are available,
the records can be validated. The first validation step regards the locality
information, which is done using the function `validateLoc()`:

```{r validateLoc, eval = TRUE, collapse = FALSE}
occs <- validateLoc(occs)
```

#### Geographical coordinates

The second validation step regards the geographical coordinates of the records,
which is performed using the function `validateCoord()`:

```{r validateCoord, eval = FALSE, collapse = FALSE}
occs <- validateCoord(occs)
``` 


#### Species taxonomy and identification

The next validation step regards the confidence level in the species
identification, which is one of the main __plantR__ features and executed by
function `validateTax()`:

```{r validateTax, eval = TRUE, collapse = FALSE}
occs <- validateTax(occs)
```

Note that the function returns up to 10 names of determiners not taken as
specialists of the family. The argument `miss.taxonomist` can be used to include
missing names of taxonomists (e.g. `miss.taxonomist = c("Arecaceae_Reis, A.")`)

#### Duplicate specimens

Another main feature of __plantR__ is the search for duplicated records across
herbaria (i.e., same biological specimen with accession numbers in two or more
collections). It uses different combinations of search strings to find direct
and indirect links between records. Besides the search itself, the user can also
homogenize information within groups of duplicates, such as species names or
geographical coordinates. This tool is performed using the function
`validateDup()`:


```{r validateDup, eval = TRUE, include=TRUE, collapse = TRUE} 
occs <- validateDup(occs)
```

### Data summary and export

Once the editing and validation steps are finished, __plantR__ provides tools
for summarizing the occurrence data, using the function `summaryData()`. In this
example, the taxonomical summary is quite uninformative, since we have only one
species.

```{r summaryData, eval = TRUE, include=TRUE, collapse= FALSE} 
summ <- summaryData(occs)
```


__plantR__ also provides an overview of the validation results (function
`summaryFlags()`):

```{r summaryFlags, eval = TRUE, include=TRUE, collapse= FALSE} 
flags <- summaryFlags(occs)
```


The package __plantR__ can also build species check-lists with vouchers using
the function `checkList()`:

```{r, eval = TRUE, collapse= FALSE, collapse = FALSE} 
checkList(occs, n.vouch = 3, type = "short")
```

Finally, __plantR__ exports data into a local folder, using function
`saveData()`, which can be used to save compressed '.csv' files based on
different grouping fields (e.g. botanical family, country, biological
collection). The export is performed using function `fwrite()` from package
__data.table__ which is quite fast even for large datasets.

<br/><br/>

# Citation

If you use this package, please cite it as: 

Lima, R.A.F et al. (2021) plantR: An R package and workflow for managing
  species records from biological collections. In preparation.


If you use the function `prepSpecies()`, please also cite the following packages
(depending on the database used):

Carvalho, G. (2020) flora: Tools for Interacting with the Brazilian Flora 2020.
R package version 0.3.4. https://CRAN.R-project.org/package=flora

Cayuela, L., Macarro, I., Stein, A. and Oksanen, J. (2017) Taxonstand: Taxonomic
Standardization of Plant Species Names. R package version 2.2.
https://CRAN.R-project.org/package=Taxonstand

If you use the function `rgbif2()`, please also cite the following package:

Chamberlain, S. et al. (2021) rgbif: Interface to the Global Biodiversity
Information Facility API. R package version 2.5.2.
https://CRAN.R-project.org/package=rgbif>.

