---
title: "Downloading speciesLink data from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{speciesLink}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = "#>",
                      eval = TRUE)
```

Installing and loading the package:

`devtools::install_github("limaRAF/plantR")`

```{r setup}
#library(rocc)
devtools::load_all() # for development
```

This package downloads information from the [speciesLink API](http://api.splink.org.br/). It generates the desired url and uses functions from `jsonlite` package to GET the url and save the output as a csv file. The speciesLink API is a courtesy of Sidnei de Souza from [CRIA](http://www.cria.org.br/) (Centro de Referência em Informação Ambiental) :)


See `?rspeciesLink` for all search options. Bellow function's arguments. 

```{r args}
args(rspeciesLink)
```


## Example 1: basic search

Search for records of *Eugenia platyphylla* and *Chaetocalyx acutifolia* in speciesLink API. Same as: [http://api.splink.org.br/records/ScientificName/Eugenia%20platyphylla/Chaetocalyx%20acutifolia/scope/plants](http://api.splink.org.br/records/ScientificName/Eugenia%20platyphylla/Chaetocalyx%20acutifolia/scope/plants)

```{r sp}
sp1 <- "Eugenia platyphylla"
sp2 <- "Chaetocalyx acutifolia"
```

Setting scope `"plants"`. 

```{r ex01}
ex01 <- rspeciesLink(filename = "ex01",
                     scientificName =  c(sp1, sp2),
                     Scope = "plants")
```

Checking search output. 

```{r ex01-check01}
head(ex01$data)
dim(ex01$data)
str(ex01$data)
```

Checking if required species are in the output. 

```{r ex01-check02}
# are required species in the output
## list species in the output
ex01.sp <- unique(ex01$data$scientificName)
## check if species are in the output - yes!
c(sp1, sp2) %in% ex01.sp
```

Checking colnames in data output.

```{r ex01-check03, eval=FALSE}
names(ex01$data)
```

```{r colnames, echo=FALSE}
knitr::kable(data.frame(columns = sort(names(ex01$data))))
```

## Example 2: specifying collection of origin and specimens with image

Search for *Rauvolfia selowii* and *Cantinoa althaeifolia*. Now using `collectioncode` and `Images` arguments.

Same as: [http://api.splink.org.br/records/CollectionCode/uec/scientificname/Rauvolfia%20sellowii/Cantinoa%20althaeifolia/Images/yes](http://api.splink.org.br/records/CollectionCode/uec/scientificname/Rauvolfia%20sellowii/Cantinoa%20althaeifolia/Images/yes)

```{r ex02}
ex02 <- rspeciesLink(filename = "ex02",
                     collectionCode = "uec",
                     scientificName = c("Rauvolfia sellowii", 
                                        "Cantinoa althaeifolia"),
                     Images = "Yes")
```

Checking again if species are in the search. 

```{r ex02-check01}
# again species are in the output
c("Rauvolfia sellowii", "Cantinoa althaeifolia") %in% ex02$data$scientificName
```

Checking url used in the search. 

```{r ex02-check02}
ex02$url
# check output
head(ex02$data)
dim(ex02$data)
str(ex02$data)
```

Is data only from UEC collection?

```{r ex02-check03}
# check fiel collectioncode
unique(ex02$data$collectionCode)
```

## Example 3: testing coordinates quality selection

For species *Tillandsia stricta*. 

```{r ex03}
ex03 <- rspeciesLink(filename = "ex03",
                     scientificName = "Tillandsia stricta",
                     Coordinates = "Yes",
                     CoordinatesQuality = "Good")
```

Checking if species is in the output.

```{r ex03-check01}
"Tillandsia stricta"%in%ex03$data$scientificName
```

Checking url and output.

```{r ex03-check02}
ex03$url
# output check
dim(ex03$data) # 1623
head(ex03$data)
```

Now with another selection of coordinate quality. 

```{r ex03b}
# another selection of coordinates quality
ex03b <- rspeciesLink(filename = "ex03b",
                      scientificName = "Tillandsia stricta")
                   #coordinatesQuality = "Bad")
```

Checking url and output.

```{r ex03b-check01}
ex03b$url
# check output
dim(ex03b$data) # 1762
head(ex03b$data)
```

## Example 4: Only plant species in IUCN Red List in a particular geographic area

This example searches for 100 herbarium plants collected in Mariana county (Minas Gerais state, Brazil) that are in the IUCN Red List. 

```{r ex04}
ex04 <- rspeciesLink(filename = "ex04",
                     Scope="plants", 
                     basisOfRecord = "PreservedSpecimen",
                     county="Mariana", 
                     stateProvince = c("Minas Gerais", "MG"),
                     country=c("Brazil", "Brasil"),
                     RedList=TRUE,
                     MaxRecords = 100)
```

```{r ex04-check01}
dim(ex04$data)
```


## Example 5: 

Checks for synonyms on [Flora do Brasil 2020](http://floradobrasil.jbrj.gov.br/reflora/listaBrasil/PrincipalUC/PrincipalUC.do;jsessionid=4887DC37EAB2ECF4A6754924CFD60AFB#CondicaoTaxonCP), the official plant list for taxonomic nomenclature.  

```{r ex05}
ex05 <- rspeciesLink(filename = "ex05",
                     scientificName = "Tillandsia stricta", 
                     Synonyms = "flora2020")
```

There is a limit of nine species for requiring to check for synonyms.  If the lenght of species name vector is > 9, function stops.

```{r ninesp}
tensp <- c(
  "Tillandsia stricta", "Rauvolfia sellowii", "Cantinoa althaeifolia",
  "Eugenia platyphylla","Chaetocalyx acutifolia", "Asplenium serra", 
  "Asplenium truncorum", "Prepusa montana", "Ocotea catharinensis",
  "Asplenium triquetrum"
            )
```

Function stops if argument `Synonyms` is used with more than nine species. If 

```{r ex05b, eval = FALSE}
ex05 <- rspeciesLink(filename = "ex05",
                     scientificName = tensp, 
                     Synonyms = "flora2020")
```

Returns error:

`Function does not support synonym check of more than nine species`

## General check 

### Checking if files were written on disk

Listing files on list. 

```{r check-resu}
resu <- list.files("results", pattern = '.csv', full.names = TRUE)
```

All outputs are readable.

```{r read-resu}
all.resu <- lapply(resu, read.csv)
```


