---
title: "Using __plantR__ to Manage Taxon Names"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: 
  - Renato A. Ferreira de Lima^[Departamento de Ciências Biológicas, ESALQ, Universidade de São Paulo, https://github.com/LimaRAF]

output: 
    rmarkdown::html_vignette:
        toc: true
        number_sections: true
    md_document:
      variant: gfm
    pdf_document:
      toc: true
      number_sections: yes
      template: null  
      toc_depth: 3
      df_print: null 
vignette: >
  %\VignetteIndexEntry{Using __plantR__ to Manage Taxon Names}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
geometry: margin=3cm
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment  = "#>",
                      echo = TRUE,
                      class.output  = "shadebox")
```

```{r load, echo = FALSE, eval = TRUE, include=FALSE}
devtools::load_all() # for development
```


<!-- ```{r setup} -->
<!-- library(plantR) -->
<!-- ``` -->

<br/><br/>

# Introduction

One of the most important steps in using information from biological
datasets is the management of taxon names. They need to be correctly
spelled and we should use only their currently accepted names based on
a given taxonomic backbone (Lima et al. in prep.). However, there are
many small corrections and standardizations that need to be made
before one can cross-validate taxonomic nomenclature, which becomes
quite burdensome as the size of the dataset increases.

__plantR__ provides tools to format, spell-check and validate taxon
names using any taxonomic backbone. The default backbone used
by __plantR__ is the [Flora e Funga do Brasil](https://floradobrasil.jbrj.gov.br/consulta). 
However, any taxonomic backbone can be used, as long as it has a
specific content and format. The companion R package __plantRdata__
provides other backbones already in this specific format from the
[World Flora Online](https://www.worldfloraonline.org/), the 
[World Checklist of Vascular Plants](https://powo.science.kew.org/), 
[GBIF](https://www.gbif.org/) and the [Taxonomic Catalog of the Brazilian Fauna](https://fauna.jbrj.gov.br)

The management of taxonomy in __plantR__ can be simply done using the
wrapper function `formatTax()` (see the section [Brief code summary] below). 
But first, we will explain in detail each step and function to manage
taxonomy within __plantR__. These functions can be applied
individually, as we will show below, but the management of taxonomic
nomenclature is potentialised when they are executed in a specific
order.

<br/><br/>

# Installing __plantR__

The package __plantR__ can be installed and loaded from [GitHub](https://github.com/LimaRAF/plantR) 
with the following codes (this may take a minute or two):
```{r, eval = FALSE}
if (!requireNamespace("remotes")) 
  install.packages("remotes")
library("remotes")

if (!requireNamespace("plantR")) 
  install_github("LimaRAF/plantR")
library("plantR")
```

For this tutorial, we will also need the global taxonomic backbones 
available from the companion R package __plantRdata__. Installing this 
package takes a couple of minutes as well, as it stores some large 
files (>10 MB).
```{r, eval = FALSE}
install_github("LimaRAF/plantRdata")
```

If you run into problems while installing __plantRdata__, remember
that it just stores the large files that we need for this tutorial.
So, you can simply download and load them into R using:

```{r, eval = FALSE}
temp_file <- tempfile()
download.file("https://raw.github.com/LimaRAF/plantRdata/refs/heads/master/data/wfoNames.rda", destfile = temp_file)
load(temp_file)
```

The code above is specific to download the World Flora Online
taxonomic backbone. You need to change the file name to download the other
backbones available [here](https://github.com/LimaRAF/plantRdata/tree/master/data).

<br/><br/>

# A practical example

Let´s create a list of names with common issues in biological
datasets. This list includes issues related to format, casing,
spelling, synonyms, typos, names with/without authors, incomplete
identifications, etc.


```{r names, eval = TRUE, collapse = FALSE}
names <- c(
  "Lindsaea sp.", "Lindsaeaceae sp.",
  "Lindsaea lancea", "Lindsaea lancia", "Lindsaea pumila",
  "Lindsaea lancea (L.) Bedd.",
  "lindsaea lancea", "Lindsaea Lancea", "LINDSAEA LANCEA",
  "Lindsaea lancea var. Falcata",
  "Lindsaea lancea var falcata",
  "Lindsaea Aff. lancea", 
  "Lindsaea Aff.lancea",
  "Lindsaea aff. lancea (L.) Bedd.",
  "Lindsaea ×improvisa K.U.Kramer",
  "Parablechnum C.Presl",
  "Blechnum spannagelii Rosenst.",
  "Blechnum cf. spannagelii", "Blechnumcf.spannagelii",
  "Blechnum austrobrasilianum de la Sota",
  "Casearia sylvestris var. angustifolia",
  "Casearia sylvestris var. angustifolia Uittien",
  "Casearia sylvestris angustifolia Uittien",
  "Casearia sylvestris Sw. var. sylvestris",
  "Blechnaceae1", 
  "Blechnum sp.2", "Blechnum sp. 2", "Blechnum sp 2", "Blechnum sp",
  "indet", "Indeterminada1"
)

```

<br/><br/>

## Preparing names using `fixSpecies()`

This function accepts both a vector of names or a data frame. In the
latter case, the data frame should contain the taxon names in a column
called, by default, 'scientificName'. The name of this column can be
defined by the user through the argument `tax.name`. And if taxon
names and authorities are available in separate columns, which is
advised, please make sure that author names are stored in a column
called 'scientificNameAuthorship' or alter this column name using the
argument `author.name`.

Here we provide to the function the vector of names defined above:

```{r fixSpecies, eval = TRUE, collapse = FALSE}
names_fixed <- fixSpecies(names)
head(names_fixed[,-c(2,4)], 7)
```

The output of `fixSpecies()` is a data frame that contains the
columns necessary for the name validation step below. For each name,
it is returned a new column containing a suggestion of a more
standardised name (i.e. 'scientificName.new'), with the isolation of
the name modifiers (e.g. cf. or aff.). Also, if the name contains
author names (and if those author names are in a detectable format),
they are split into two different columns ('scientificName.new' and
'scientificNameAuthorship.new').

The new column 'scientificNameStatus' stores all possible flags
detected in the original name. These flags include the detection and
standardization of open nomenclature (e.g. aff., cf.), infra-specific
levels (e.g. var., subsp., f.), hybrids, incomplete identifications
and undeterminations. It also flags and solves issues related to name
casing and the notation of morphotypes.

### Internal functions

Besides the flagging of particular cases of taxon names, the function
`fixSpecies()` contains different internal functions that standardise
the:

* notation of open nomenclature abbreviations and name modifiers (function `fixAnnotation()`);
* notation of incomplete identifications and undeterminations (function `fixIndet()`);
* the casing of taxon names (function `fixCase()`);
* the separation between taxon names and taxon name authorships (function `fixAuthors()`).

These internal functions may also be useful in itself depending on the
user's goals. Below, some examples of their isolate use:

```{r fixSpecies_internals, eval = TRUE, collapse = TRUE}
plantR:::fixAnnotation(c("Lindsaea lancea var falcata", "Lindsaea Aff.lancea"))
plantR:::fixIndet(c("Indet1", "Blechnum sp. 2", "Blechnum sp 2", "Blechnum sp"))
plantR:::fixCase(c("lindsaea lancea", "Lindsaea Lancea", "LINDSAEA LANCEA"))
plantR:::fixAuthors(c("Lindsaea lancea (L.) Bedd.", "Parablechnum C.Presl"))
```

<br/><br/>

## Validating taxon names using `prepSpecies()`

After the standardization of notation and format of taxon names, it is
possible to validate them against a taxonomic backbone, aiming at
finding synonyms, orthographic variants and/or typos.

In __plantR__ this validation of the taxonomic nomenclature is done
using the function `prepSpecies()`: 

```{r prepSpecies, eval = TRUE, collapse = TRUE}
names_valid <- prepSpecies(names_fixed,
                           tax.names = c("scientificName.new", 
                                         "scientificNameAuthorship.new"))
head(names_valid[,-c(2,3,4,9,11)], 7)
```

The output of the function is optimized to ease the user´s interpretation
and decision-making. The output column 'tax.notes' contains the
important comments related to the name validation, while the argument
`drop.cols` controls which columns should be returned (set 
`drop.cols = ""` to get all columns). By default, synonyms and
orthographic variants are replaced by the name accepted in the
reference backbone.

In addition, `prepSpecies()` has arguments to control the minimum
similarity allowed between names in fuzzy matching (argument `sug.dist`)
and the editing and cleaning of names before matching (arguments
`clean.indet` and `clean.names`). 

The function was conceived to validate thousands of names at once.
So, to speed up the validation process, users can validate names by
their initial letter (argument `split.letters`) or parallelise
the computation (arguments `parallel` and `cores`).

By default, `prepSpecies()` uses the internal backbone from the 
[Flora e Funga do Brasil](https://floradobrasil.jbrj.gov.br/consulta).
But any backbone in the right format can be provided by the user via
the argument `db` of `prepSpecies()`. The __plantRdata__ package,
available only on GitHub at [this link](https://github.com/LimaRAF/plantRdata), 
provides preformatted objects with backbones from different sources. 
From __plantRdata__ we are going to use two other taxonomic backbones 
from the [World Flora Online](https://www.worldfloraonline.org/) and 
the [World Checklist of Vascular Plants](https://powo.science.kew.org/) 
(use `?wfoNames` and `?wcvpNames` for details on those backbones).

```{r prepSpecies_other, eval = TRUE, collapse = TRUE}
# loading the WFO and WCVP backbones into a temporary environment
temp.env <- new.env(parent = emptyenv())
data(list = c("wfoNames", "wcvpNames"), package = "plantRdata", 
     envir = temp.env)
# using the World Flora Online
names_valid_wfo <- prepSpecies(names_fixed,
                               tax.names = c("scientificName.new", 
                                         "scientificNameAuthorship.new"),
                               db = temp.env$wfoNames)
# using the World Checklist of Vascular Plants
names_valid_wcvp <- prepSpecies(names_fixed,
                                tax.names = c("scientificName.new", 
                                         "scientificNameAuthorship.new"),
                                db = temp.env$wcvpNames)
# Comparing the results
names_bfo_wfo_wcvp <- cbind.data.frame(names_valid$scientificName.new,
                                  names_valid$scientificNameFull,
                                  names_valid_wfo$scientificNameFull,
                                  names_valid_wcvp$scientificNameFull)
diff <- names_valid$scientificNameFull != names_valid_wfo$scientificNameFull
diff[is.na(diff)] <- FALSE
head(names_bfo_wfo_wcvp[diff, ], 3)
```

Note that the computing speed when using larger backbones (over a
million names) is very different than when using smaller ones (a few
hundred thousand names). This is because the name-matching process is
done pairwise between the input and reference names. But please try to
play with the arguments `split.letters` and `parallel` and `cores` to
investigate how fast it can be.

Note as well that each taxonomic backbone provides a different opinion
on each name. So it is up to the user to select the best backbone for
its own purposes.

### Internal functions

The most important internal function used by `prepSpecies()` is
`nameMatching()`, which is the function that actually performs the
exact and fuzzy matching between the input names against the reference
names from the taxonomic backbone selected. As mentioned above, the
name matching can be parallelized and/or performed separately by initial
letters to speed up computational time for larger datasets. But
basically, it compares a set of input and reference names trying to
find exact or fuzzy matches:

```{r prepSpecies_internals, eval = TRUE, collapse = TRUE}
input_names <- c("Casearia silvestris", "Casearia decandra")
ref_names <- c("Casearia aculeata", "Casearia arborea", 
               "Casearia decandra", "Casearia sylvestris")
nameMatching(input_names, ref_names)
```

<br/><br/>

## Validating family names using `prepFamily()`

__plantR__ contains an internal dictionary of valid family names which 
can be used via the function `prepFamily()`. Currently, valid family
names are based on the APG IV (angiosperms) and the PPG I (ferns
and lycophytes). And if the family info is missing it can also be found 
based on genus information.

```{r prepFamily, eval = TRUE, collapse = TRUE}
names_valid <- prepFamily(names_valid,
                          fam.name = "suggestedFamily",
                          spp.name = "scientificName.new")
```

It was not the case in this example, but `prepFamily()` returns warnings
in the case of conflicts between the original family names and the
family name suggested by the APG IV/PPG I family list. In this example,
`prepFamily()` just added the missing family names based on the
names of the genera, in the new column called 'family.new'.

<br/><br/>

# Brief code summary 

In this tutorial, we described many steps and details related to the
functions that manage taxonomy within __plantR__. Here we provide the
main codes that you actually need to include in your scripts, assuming
that you already created the vector of taxon names called `names` and 
that you only want to validate it against the 
[Flora e Funga do Brasil](https://floradobrasil.jbrj.gov.br/consulta):

```{r code_summary, eval = TRUE, collapse = TRUE}
names_fixed <- fixSpecies(names)
names_valid <- prepSpecies(names_fixed,
                           tax.names = c("scientificName.new", 
                                         "scientificNameAuthorship.new"))
names_valid <- prepFamily(names_valid, 
                          fam.name = "suggestedFamily",
                          spp.name = "scientificName.new")
```

Or, as mentioned above, even simpler using the wrapper `formatTax()`:

```{r summary, eval = TRUE, collapse = TRUE}
names_df <- data.frame(scientificName = names)
names_df_valid <- formatTax(names_df)
```


<br/><br/>

# Citation

If you use __plantR__, please cite it as: 

> Lima, R.A.F., Sánchez-Tapia, A., Mortara, S.R., ter Steege, H.,
Siqueira, M.F. (2023). *plantR*: An R package and workflow for
managing species records from biological collections. Methods in
Ecology and Evolution 14(2): 332-339.
https://doi.org/10.1111/2041-210X.13779

And please also cite the taxonomic backbones that you used:

> Borsch, T., Berendsohn, W., Dalcin, E., et al. (2020). World Flora
Online: Placing taxonomists at the heart of a definitive and
comprehensive global resource on the world’s plants. Taxon, 69(6):
1311-1341. https://doi.org/10.1002/tax.12373

> Govaerts, R., Nic Lughadha, E., Black, N. et al. (2021). The World
Checklist of Vascular Plants, a continuously updated resource for
exploring global plant diversity. Sci. Data 8: 215.
https://doi.org/10.1038/s41597-021-00997-6

> Flora e Funga do Brasil (Constantly updated): Flora e Funga do Brasil project. Instituto de Pesquisas Jardim Botânico do Rio de Janeiro. Dataset/Checklist. https://doi.org/10.15468/1mtkaw

<br/><br/>

# References

> Lima, R.A.F. et al. (in prep.) Ten simple rules of taxonomy for
ecological datasets.

<br/><br/>

# Bug report and suggestions

The plantR project is hosted on [GitHub](https://github.com/LimaRAF/plantR/).
Please report any bugs and suggestions for improvements to the package
[here](https://github.com/LimaRAF/plantR/issues).

